A load balancing scheme for massively multiplayer online games \thanks{This work was supported by the National Research Council (CNPq) and by the Coordination of Improvement of Higher Education (CAPES), both brazilian research funding agencies.}

### 2009-04-10 trunk/textos/dissertation/MTAP_LaTeX/pub/tex/latex/svjour3/global/balancing.tex

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2006/03/15
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallextended]{svjour3}     % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}         % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{algorithmic}
\usepackage{algorithm}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}
%[TODO:ajustar TODAS as \refs]
\title{A load balancing scheme for MMOGs %\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

@@

\institute{Carlos Eduardo Benevides Bezerra, Cláudio Fernando Resin Geyer \at
							Universidade Federal do Rio Grande do Sul \\
              Av. Bento Gonçalves, 9500, Porto Alegre, Brazil \\
              Tel.: +55-51-84065986\\
              Fax.: +55-51-33087308\\
              \email{carlos.bezerra@inf.ufrgs.br, geyer@inf.ufrgs.br}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
%           \and
%           Cláudio Fernando Resin Geyer \at
%           		Universidade Federal do Rio Grande do Sul \\
%              Av. Bento Gonçalves, 9500, Porto Alegre, Brazil \\
%              Tel.: +55-51-99011155\\
%              \email{geyer@inf.ufrgs.br}
}

@@

The main characteristic of massively multiplayer games is the large number of players, having dozens, or even hundreds, of thousands participants simultaneously \cite{schiele2007rpp}. This large number of players interacting with one another generates a traffic on the support network which grows quadratically compared to the number of players, in the worst case.

When using a client-server architecture, it is necessary that the server intermediates the communication between each pair of players -- assuming that the game is intended to provide guarantees of consistency and resistance to cheating. Obviously, this server will have a large communication load, thus, it must have enough resources (available bandwidth) to meet the demand of the game. We consider here that the main resource to analyse is the available bandwidht, for this is the current bottleneck for MMOGs \cite{feng2007wnn}.

The question is, when you use a distributed server, you must delegate to each server node a load proportional to its power. Thus, no matter to which server each player is connected, their game experience will be similar, regarding the response time for their actions and the time it takes to be notified of actions from other players as well as of state changes in the virtual environment of the game.

An initial idea might be to distribute the players between servers, so that the number of players on each server would be proportional to the bandwidth of that server. However, this distribution would not work, for the burden caused by players also depends on how they are interacting with one another. For example, if the avatars of two players are too distant from each other, probably there will be no interaction between them and therefore the server needs only to update each one of them with the result of their own actions. However, if these avatars are close to each other, each player should be updated not only of his own actions, but also of the actions of the other player.

@@

For this reason, it is not enough just to divide the players between servers, even if this division is proportional to the resources of each one of them. First, in some cases the usage of the server's bandwidth is square to the number of players, while in others it is proportional. That reason alone is enough to get another criterion for load balancing. Moreover, there is another important issue: the existence of hotspots, which motivates the creation of a load balancing scheme for MMOGs to prevent that the presence of hotspots degrade the quality of the game beyond the tolerable.

%[TODO:falar das próxima seções]

@@

It is perceived, then, that each state will be sent twice for each pair of players who communicate through different servers. This overhead not only cause the waste of resources of the servers, but it also increases the delay to update the status of the replica of the game in the players' machines, damaging the interaction between them.

Therefore, players who are interacting with each other should, ideally, be connected to the same server. However, it is possible that all players are linked through relations of interaction. For example, two avatars of two different players, may be distant from each other, but both could be interacting with a third avatar, between the other two. However, it is still necessary to divide them between servers. The question is how many pairs of players and which of them will be divided into different server nodes. It is therefore necessary to decide a criterion to group players.

%[TODO: figura da interação de A, B e C]

Next, some principles, used by other authors, will be presented. Based on some of these principles, and on research and implementation carried out, it will be defined the load balancing scheme proposed here.

@@

%O balanceamento de carga entre servidores de jogos online maciçamente multijogador é fortemente dependente da distribuição dos avatares dos jogadores através do ambiente virtual. Além disso, a depender da concentração de avatares, pode-se alternar entre uma função de crescimento de tráfego linear e uma função quadrática. Sendo assim, é necessário lidar com a localidade dos avatares, de maneira a otimizar o uso de largura de banda do sistema servidor, minimizando, tanto quanto possível, o overhead causado pela comunicação entre jogadores ligados a servidores diferentes.

One way to use the locality of the players is grouping them according to the position occupied by their avatars in the virtual environment. The question would be how to form such groups. One way of doing this would be dividing the world of the game in several cells connected to one another. Each cell would be a part of the world, with its own content and characteristics, which would be assigned to a server node. The easiest way to do this is with a grid of cells of the same size and shape.

%[TODO:footnote?]Porém, o formato e a disposição destas células irá influenciar no tráfego gerado pelas mesmas entre os servidores. Por exemplo, um ambiente bidimensional poderia ser dividido em uma grade de células quadradas. Neste caso, cada uma destas células teria oito vizinhos, em média -- células nas bordas do mapa poderiam ter cinco ou três vizinhos apenas. Quanto mais servidores vizinhos, maior o tráfego entre servidores, e maior o overhead causado por esta comunicação. A distribuição ideal, então, seria utilizando células hexagonais, cada uma com seis vizinhos. Estudos comprovam que esta é a divisão em células iguais que permite o menor número de vizinhos por célula [TODO:ref]. Outra possibilidade seria utilizando fileiras alternadas de células quadradas, onde cada fileira seria deslocada o equivalente à metade do comprimento de uma célula.

%[TODO:figura com diferentes tipos de divisão:sqr/hex/shift]

@@

%[TODO: tirar todas as contrações!]
Obviously, the microcells designated to the same server node do not generate additional traffic to synchronize with each other, but the synchronization overhead of the macrocell is unpredictable, because the number of neighbors of each one of them is not foreknown, for its shape is variable. However, it was demonstrated \cite{devleeschauwer2005dma} that this overhead is compensated by a better distribution of the load between servers in the game. %Figure [TODO: ref] illustrates the division of a virtual environment in two-dimensional microcells.
%It is illustrated in [TODO: ref] in the grouping of these micro macrocélulas dynamics, which can adapt to the distribution of avatars.

%[TODO: figuraS com microcélulas e macrocélulas - fazer vetorial ou copiar? depende do tempo que tiver depois de terminar o texto. divisao em micro e agrupamento em macro]

@@

In \cite{lee2003sdl}, it is also proposed a scheme for dynamic load balancing for the servers of a multi-server virtual environment, taking into consideration that users can be distributed through this world in a non-uniform way. Following the scheme proposed by the authors, an overloaded server starts the process by selecting a number of other servers to be part of the load redistribution. The set of selected servers depend on the load level of the initiating server as well as on the amount of idle resources of the other servers. After the formation of this set, its elements allocate portions of the virtual environment using a graph partitioning algorithm, so that the servers involved have similar final load.

The main aspect of the solution proposed by the authors was the use of local information (the server that initiated the balancing process and its neighbors), rather than global information (involving all servers in the balance). The former shows small overhead, but it can not solve the problem efficiently in a few steps, as overloaded servers tend to be adjacent. The global approach, in turn, is able to divide the workload in the most balanced way possible, but its complexity grows exponentially with the number of servers involved. The solution is then by involving only a subset of servers, such that its cardinality varies according to the need (if the neighbors of the server which triggered the load balancing are also overburdened, more servers are selected) . Thus, there is a little more information than the local approach, but without the complexity inherent to the global one.

%**	
%To tackle the problem, the authors also subdivide the virtual environment in rectangular \textbf{cells} -- similar to the microcrells. The cells are grouped in \textbf{regions} -- or macrocells -- and each region is managed by one server. Each server keeps its players updated of the state of the virtual environment, while also handling the interactions between avatars in its region. Two cells are adjacent (or neighbors) if they share a border. Similarly, two regions -- and their servers -- are adjacent if there is a pair of adjacent cells, each of which belong to one of the two regions. Figure \ref{fig:leeA} illustrates a distributed server system, consisting of 16 nodes. The virtual environment is divided in 256 cells, which are grouped in 16 regions.

%Para atacar o problema, os autores também subdividem o ambiente virtual em \textbf{células} -- semelhantes às microcélulas -- retangulares, sendo que o número de servidores é muito menor que o número de células. As células são agrupadas em \textbf{regiões} -- ou macrocélulas -- e cada região é gerenciada por um servidor. Cada servidor mantém atualizadas as informações de estado dos usuários e lida com as interações entre avatares na região a ele dedicada. Cada usuário envia e recebe atualizações de estado através do servidor que gerencia a região na qual ele está jogando. Duas células são ditas adjacentes (ou vizinhas) se elas compartilharem uma fronteira. Analogamente, duas regiões, e seus respectivos servidores, são ditos adjacentes se existir um par de células adjacentes, cada uma das quais pertencendo a uma das duas regiões.

To tackle the problem, the authors also subdivided the virtual environment in rectangular \textbf{cells} -- similar to the microcells --, where the number of servers is much smaller than the number of cells. The cells are grouped into \textbf{regions} -- or macrocells -- and each region is managed by one server (Figure \ref{fig:macromicro}. Each server handles the interaction between avatars in the region dedicated to him. It receives the inputs of the players controlling these avatars and sends back to them the up-to-date game state. Two cells are called adjacent (or neighbors) if they share a border. Similarly, two regions, and the servers assigned to them, are called adjacent (neighbors) if there is a pair of adjacent cells, each of which belonging to one of the two regions.

\begin{figure}
 \centering
 \includegraphics[width=0.4\textwidth]{images/macromicro}
 \caption{Cells grouped into four regions (R$_\text{1}$, R$_\text{2}$, R$_\text{3}$ and R$_\text{4}$)}
 \label{fig:macromicro}
\end{figure}

%**
%\begin{figure}
% \centering
% \includegraphics[width=0.8\textwidth]{images/LEE_leeA.jpg}
% \caption{Cells and regions in a distributed server system}
% \label{fig:leeA}
%\end{figure}

%**
The workload of a cell was defined as the number of avatars present in that cell. The authors assumed that all players receive state updates in the same frequency, so that the burden of processing (computing and communication) that a cell requires from a server is proportional to the number of users in that cell. The workload of a region and its designated server is defined as the sum of the individual workloads of the cells which form the region. Each server periodically evaluates its workload and exchange this information with its neighbors. They have assumed, too, that these servers are connected through a high-speed network. Thus, the overhead to exchange workload information among neighbors is limited and considered negligible compared to other costs of the distribution. For the same reason, they also assumed as negligible the overhead of communication between servers in different regions when players are interacting.

\subsubsection{Selection of local server group to balance the load}
\label{sec:regselect}

A server starts the balancing when the load assigned to it is beyond its power. This server selects a number of other servers to get involved with the distribution. First, it chooses the least loaded server among its neighbors and sends a request that he participates in the load balancing. The chosen server rejects the request if it is already involved in another balancing group, otherwise it responds to the server with the load information of its own neighbors. If the selected neighbor server is unable to absorb the extra workload of the initiating server, the selection is performed again among the neighbors not only of the overloaded server, but also the neighbors of the neighbor chosen in the first step. The selection continues until the workload of the first server can be absorbed -- that is, the workload of all selected servers becomes smaller than a certain limit.

Figure \ref{fig:lee2} illustrates the operation of the algorithm. All servers have the same capacity, each one being able to handle 100 users. First, the initiator server, $S_6$ is inserted into SELECTED and its neighbors ($S_2, S_5, S_7$ and $S_{10}$) are added to CANDIDATES (Figure \ref{fig:lee2}(a)). So $S_7$, which has the lowest workload among the servers in CANDIDATES, is selected and invited to participate in the load distribution. When $S_7$ sends to $S_6$ the workload information of its neighbors ($S_3, S_6, S_8$ and $S_{11}$), $S_7$ is inserted into SELECTED and its neighbors, except $S_6$, are added to CANDIDATES (Figure \ref{fig:lee2}(b)). Now, $S_{11}$, which has the lowest workload among servers in CANDIDATES, is selected and invited to participate in the load distribution. However, $S_{11}$ rejects the invitation, because it is involved in another distribution, initiated by $S_{12}$. Thus, $S_{11}$ is removed from CANDIDATES and $S_{10}$ is selected because it now has the lowest workload among all servers in CANDIDATES (Figure \ref{fig:lee2}(c)). This process continues until the average workload is under a pre-defined threshold (Figure \ref{fig:lee2}(d) and Figure \ref{fig:lee2}(e)).

%[TODO:pegar a figura em ingles]

@@

%Um problema clássico de alocação de tarefas em sistemas distribuídos é o da dependência entre elas [TODO:refsDEdistGRAPH]. Tarefas dependentes entre si fazem com que os processadores nos quais elas estão sendo executadas tenham que se comunicar para que o processamento possa continuar. Isso gera dois problemas principais: em primeiro lugar, o processamento como um todo é atrasado por causa do tempo de espera de cada transmissão e recebimento de mensagens -- supondo que a comunicação entre os processos seja por meio de mensagens, como é o caso do MPI [TODO:ref].

@@

%Para resolver este problema, são utilizados grafos, da seguinte maneira: o conjunto de tarefas a realizar é mapeado em um grafo com pesos. Cada vértice representa uma tarefa e cada aresta representa a comunicação entre as tarefas. O peso de cada vértice representa o custo de processamento e o peso da aresta a carga de comunicação. Para fazer a distribuição, é feito um particionamento do grafo gerado. Cada partição terá um conjunto de tarefas que serão executadas no mesmo nodo e, conseqüentemente, o atraso de comunicação entre elas será relativamente pequeno [TODO:refbli]. No entanto, as arestas do grafo que ligam partições diferentes representam comunicação entre diferentes nodos. A Figura [TODO:ref] ilustra essa situação.

@@

%[TODO:figura do grafo de tarefas e dependências, assim como particionamento com corte de arestas e diferentes particionamentos levando a diferentes cortes de aresta]

@@

%Os algoritmos de particionamento do grafo de tarefas, geralmente, buscam atingir dois objetivos: gerar partições de peso aproximadamente igual, ou seja, cada nodo terá uma carga computacional semelhante à dos outros; e minimizar o corte de aresta, ou seja, fazer com que a dependência entre os conjuntos de tarefas seja a menor possível, reduzindo a comunicação entre os nodos de processamento. Por corte de aresta, entende-se o somatório dos pesos de todas as arestas que ligam vértices que estão em partições diferentes. [TODO:reffig]

@@

%%TODO:ver esse "definições e parametros"
%Nas próximas seções, será apresentada a solução proposta neste trabalho para balanceamento de carga, começando pelas definições e parâmetros utilizados, assim como será descrito como o problema foi mapeado para grafos e quais os objetivos e critérios do esquema proposto.

@@

	\item \textbf{System usage}: fraction that indicates how much of system resources as a whole is being used. It is defined by:

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2006/03/15
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallextended]{svjour3}     % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}         % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{algorithmic}
\usepackage{algorithm}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}
%[TODO:ajustar TODAS as \refs]
\title{A load balancing scheme for massively multiplayer online games \thanks{This work was supported by the National Research Council (CNPq) and by the Coordination of Improvement of Higher Education (CAPES), both brazilian research funding agencies.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

@@

\institute{Carlos Eduardo Benevides Bezerra, Cláudio Fernando Resin Geyer \at
							Universidade Federal do Rio Grande do Sul \\
              Av. Bento Gonçalves, 9500, Porto Alegre, Brazil \\
              Tel.: +55-51-84065986\\
              Fax.: +55-51-33087308\\
              \email{carlos.bezerra@inf.ufrgs.br, geyer@inf.ufrgs.br}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
%           \and
%           Clï¿½udio Fernando Resin Geyer \at
%           		Universidade Federal do Rio Grande do Sul \\
%              Av. Bento Gonï¿½alves, 9500, Porto Alegre, Brazil \\
%              Tel.: +55-51-99011155\\
%              \email{geyer@inf.ufrgs.br}
}

@@

The main characteristic of massively multiplayer games is the large number of players, having dozens, or even hundreds, of thousands of participants simultaneously \cite{schiele2007rpp}. This large number of players interacting with one another generates a traffic on the support network which may grow quadratically compared to the number of players, in the worst case (Figure \ref{fig:quadratic}).

\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{images/quadratic}
  \caption{Quadratic growth of traffic when avatars are close to each other}
  \label{fig:quadratic}
\end{figure}

When using a client-server architecture, it is necessary that the server intermediates the communication between each pair of players -- assuming that the game is intended to provide guarantees of consistency and resistance to cheating. Obviously, this server will have a large communication load, thus, it must have enough resources (available bandwidth) to meet the demand of the game. We consider here that the main resource to analyse is the available bandwidth, for this is the current bottleneck for MMOGs \cite{feng2007wnn}.

The problem is that, when using a distributed server, it must be delegated to each server node a load proportional to its power. Thus, no matter to which server each player is connected, their game experience will be similar, regarding the response time for their actions and the time it takes to be notified of actions from other players as well as of state changes in the virtual environment of the game.

An initial idea might be to distribute the players among servers, so that the number of players on each server would be proportional to that server's bandwidth. However, this distribution would not work, for the burden caused by players also depends on how they are interacting with one another. For example, if the avatars of two players are too distant from each other, probably there will be no interaction between them and therefore the server needs only to update each one of them with the result of their own actions. However, if these avatars are close to each other, each player should be updated not only of his own actions, but also of the actions of the other player.

@@

For this reason, it is not enough just to divide the players between servers, even if this division is proportional to the resources of each one of them. First, in some cases the usage of the server's bandwidth is square to the number of players, while in others it is proportional. This reason alone is enough to define a new criterion for load balancing. Moreover, there is another important issue: the overhead of the distribution. As the servers need to communicate with one another, there must be a way to minimize this traffic, reducing the waste of resources of the server system. The load balancing scheme for MMOGs must, then, prevent the presence of hotspots from degrading the quality of the game beyond a tolerable limit.

%[TODO:falar das prï¿½xima seï¿½ï¿½es]

@@

It is perceived, then, that each state will be sent twice for each pair of players who communicate through different servers. This overhead not only causes the waste of resources of the servers, but it also increases the delay to update the status of the replica of the game in the players' machines, damaging the interaction between them.

Therefore, players who are interacting with each other should, ideally, be connected to the same server. However, it is possible that all players are linked through relations of interaction. For example, two avatars of two different players, may be distant from each other, but both could be interacting with a third avatar, between them. Anyway, it is still necessary to divide the players among servers. The question is how many pairs of players and which of them will be divided into different server nodes. It is therefore necessary to decide a criterion to group players. Many works have been done in the past -- such as \cite{devleeschauwer2005dma,lu2006lbm,chen2005lad,duong2003dls,ahmed2008mol} --  trying to find a near optimal, yet fast, load balancing technique.

In \cite{devleeschauwer2005dma}, for example, a few algorithms are proposed. Their idea is to transfer regions managed by overloaded servers to  lightly loaded ones, in a way such that the maximum overload among the servers is reduced. Some very interesting principles are used, as the concepts of microcells and macrocells (which will be described in the next section). However, their model has some differences to ours. For instance, they consider the number of players as the load to distribute, heterogeneous systems are not supported, and the algorithm they use to refine an existing partitioning is simulated annealing. This one may be a little too general, considering that there are algorithms specific to partitioning, such as \cite{kernighan1970ehp} or \cite{fiduccia1982lth}, which may perform better. Finally, some of the algorithms proposed in \cite{devleeschauwer2005dma} do not consider the communication overhead between servers.

It is proposed in \cite{lu2006lbm} a load balancing of cpu usage among the nodes of a cluster. The load balancing is done by an ``off the shelf'' NAT load balancer, which distributes the players' connections among the cluster nodes in a round-robin manner. Thus, each cluster node will end up with the same number of players, on the average. This approach, although it does balance the number of clients per cluster node, communication between the server nodes is not considered, for they are connected via a high-speed and low latency network. Furthermore, in this work we are not concerned with cpu usage, but bandwidth occupation.%OK

Chen et al. propose, in \cite{chen2005lad}, a load shedding technique, in which an overloaded server attempts do shed its load to its neighbors. After finding a lightly loaded neighbor server -- if there is one -- it transfers some of its boundary microcells to that neighbor. To form the group of microcells to transfer, a microcell from the border is chosen, and others are added in a breadth-first search (BFS) order. A very similar work, which also uses BFS, is described in \cite{duong2003dls}. However, these models present a few points which could be improved. For example, a server is considered overloaded when a certain number $N$ of players is reached. As we said before, this is not the best measure, since the load on a server may vary completely, depending on how these $N$ players are distributed across the region managed by the server. Also, the use of BFS to merge two partitions or split one of them may not be ideal. Some other algorithm for graph partitioning applied to distributed systems \cite{kernighan1970ehp,fiduccia1982lth,karypis1999fah,hendrickson1995isg} could be used, for most of these algorithms were designed exactly to balance load and minimize dependence between nodes.

Ahmed and Shirmohammadi propose a microcell oriented load balancing model \cite{ahmed2008mol}. To balance the load, their algorithm, first, finds all clusters of microcells which are managed by the overloaded server. After that, the smallest cluster (in number of microcells) is selected and then, from that cluster, it is chosen the microcell with the lowest communication with other microcells managed by the same server. The chosen microcell is then transferred to the least loaded server. This process is repeated while the server is still overloaded and there are servers which can handle the extra load of the microcells being transferred. The authors define the load on a server, as opposed to other load balancing schemes, as the summing of the message rate of every player that a server must handle, which is a much more realistic measure than the number of players on the server. Also, their model for load balancing tries to minimize the inter-server communication, improving the overall system performance. Their model will be compared to ours, for it takes into consideration many of the aspects that we consider. The comparison will be performed through some simulations, which are described in section \ref{sec:simulations}.

Next, some principles, defined in previous works, will be presented. Based on some of these principles, and on research and implementation carried out, it will be defined the load balancing scheme proposed here.

@@

%O balanceamento de carga entre servidores de jogos online maciï¿½amente multijogador ï¿½ fortemente dependente da distribuiï¿½ï¿½o dos avatares dos jogadores atravï¿½s do ambiente virtual. Alï¿½m disso, a depender da concentraï¿½ï¿½o de avatares, pode-se alternar entre uma funï¿½ï¿½o de crescimento de trï¿½fego linear e uma funï¿½ï¿½o quadrï¿½tica. Sendo assim, ï¿½ necessï¿½rio lidar com a localidade dos avatares, de maneira a otimizar o uso de largura de banda do sistema servidor, minimizando, tanto quanto possï¿½vel, o overhead causado pela comunicaï¿½ï¿½o entre jogadores ligados a servidores diferentes.

One way to explore the locality of the players is by grouping them according to the position occupied by their avatars in the virtual environment. The question would be how to form such groups. One way of doing this would be dividing the world of the game in several cells connected to one another. Each cell would be a part of the world, with its own content and characteristics, which would be assigned to a server node. The easiest way to do this is with a grid of cells of the same size and shape.

%[TODO:footnote?]Porï¿½m, o formato e a disposiï¿½ï¿½o destas cï¿½lulas irï¿½ influenciar no trï¿½fego gerado pelas mesmas entre os servidores. Por exemplo, um ambiente bidimensional poderia ser dividido em uma grade de cï¿½lulas quadradas. Neste caso, cada uma destas cï¿½lulas teria oito vizinhos, em mï¿½dia -- cï¿½lulas nas bordas do mapa poderiam ter cinco ou trï¿½s vizinhos apenas. Quanto mais servidores vizinhos, maior o trï¿½fego entre servidores, e maior o overhead causado por esta comunicaï¿½ï¿½o. A distribuiï¿½ï¿½o ideal, entï¿½o, seria utilizando cï¿½lulas hexagonais, cada uma com seis vizinhos. Estudos comprovam que esta ï¿½ a divisï¿½o em cï¿½lulas iguais que permite o menor nï¿½mero de vizinhos por cï¿½lula [TODO:ref]. Outra possibilidade seria utilizando fileiras alternadas de cï¿½lulas quadradas, onde cada fileira seria deslocada o equivalente ï¿½ metade do comprimento de uma cï¿½lula.

%[TODO:figura com diferentes tipos de divisï¿½o:sqr/hex/shift]

@@

%[TODO: tirar todas as contraï¿½ï¿½es!]
Obviously, the microcells designated to the same server node do not generate additional traffic to synchronize with each other, but the synchronization overhead of the macrocell is unpredictable, because the number of neighbors of each one of them is not foreknown, for its shape is variable. However, it was demonstrated \cite{devleeschauwer2005dma} that this overhead is compensated by a better distribution of the load between servers in the game. %Figure [TODO: ref] illustrates the division of a virtual environment in two-dimensional microcells.
%It is illustrated in [TODO: ref] in the grouping of these micro macrocï¿½lulas dynamics, which can adapt to the distribution of avatars.

%[TODO: figuraS com microcï¿½lulas e macrocï¿½lulas - fazer vetorial ou copiar? depende do tempo que tiver depois de terminar o texto. divisao em micro e agrupamento em macro]

@@

In \cite{lee2003sdl}, it is also proposed a scheme for dynamic load balancing for the servers of a multi-server virtual environment. Following the scheme proposed by the authors, an overloaded server starts the process by selecting a number of other servers to be part of the load redistribution. The set of selected servers depends on the load level of the initiating server as well as on the amount of idle resources of the other servers. After the formation of this set, its elements allocate portions of the virtual environment using a graph partitioning algorithm, so that the servers involved have similar final load.

To achieve this goal, the authors also subdivided the virtual environment in rectangular \textbf{cells} -- similar to the microcells --, where the number of servers is much smaller than the number of cells. The cells are grouped into \textbf{regions} -- or macrocells -- and each region is managed by one server (Figure \ref{fig:macromicro}). Each server handles all the interactions between avatars located in the region assigned to it. It receives the inputs of the players controlling these avatars and sends back to them the up-to-date game state.

\begin{figure}
 \centering
 \includegraphics[width=0.4\textwidth]{images/macromicro}
 \caption{Cells grouped into four regions (R$_1$, R$_2$, R$_3$ and R$_4$)}
 \label{fig:macromicro}
\end{figure}

Two cells are called adjacent (or neighbors) if they share a border. Similarly, two regions -- and the servers assigned to them -- are called adjacent (neighbors) if there is a pair of adjacent cells, each of which belonging to one of the two regions. The workload of a cell was defined as the number of avatars present in that cell. The authors assumed that all players receive state updates of the same lenght and in the same frequency, so that the burden of processing (computing and communication) that a cell requires from a server is proportional to the number of users in that cell. The workload of a region and its designated server is defined as the sum of the individual workloads of the cells which form the region. Each server periodically evaluates its workload and exchange this information with its neighbors. They have assumed, too, that these servers are connected through a high-speed network. Thus, the overhead to exchange workload information among neighbors is limited and considered negligible compared to other costs of the distribution. For the same reason, they also assumed as negligible the overhead of communication between servers in different regions when players are interacting.

The main aspect of the solution proposed by the authors was the use of local information (the server that initiated the balancing process and its neighbors), rather than global information (involving all servers in the balance). When a server is overloaded, it searches for lightly loaded servers close to it in the overlay network. A breadth-first lookup for lightly loaded neighbor servers -- as proposed in \cite{duong2003dls} -- causes a small overhead, but it may not solve the problem efficiently in a few steps, as overloaded servers tend to be adjacent. The global approach, in turn, is able to divide the workload in the most balanced way possible, but its complexity may become too high.

The solution proposed by the authors is then by involving only a subset of servers, such that its cardinality varies according to the need (if the neighbors of the server which triggered the load balancing are also overburdened, more servers are selected). However, this set is formed in a more clever way than by simply using a breadth-first search. The algorithm used to find the servers is described in the next section.

\subsubsection{Selection of a local server group to balance the load}
\label{sec:regselect}

A server starts the balancing when the load assigned to it is beyond its capacity. This server selects a number of other servers to get involved with the distribution. First, it chooses the least loaded server among its neighbors and sends a request that he participates in the load balancing. The chosen server rejects the request if it is already involved in another balancing group, otherwise it responds to the server with the load information of its own neighbors. If the selected neighbor server is unable to absorb all the extra workload of the initiating server, the selection is performed again among the neighbors not only of the overloaded server, but also the neighbors of the already selected servers. The selection continues until the workload of the first server can be absorbed -- that is, the workload of all selected servers becomes smaller than a certain limit.

Figure \ref{fig:lee2} illustrates the operation of the algorithm. All servers have the same capacity, each one being able to handle 100 users. First, the initiator server, $S_6$ is inserted into SELECTED and its neighbors ($S_2, S_5, S_7$ and $S_{10}$) are added to CANDIDATES (Figure \ref{fig:lee2}(a)). So $S_7$, which has the lowest workload among the servers in CANDIDATES, is selected and invited to participate in the load distribution. When $S_7$ sends to $S_6$ the workload information of its neighbors ($S_3, S_6, S_8$ and $S_{11}$), it is inserted into SELECTED and its neighbors, except $S_6$, are added to CANDIDATES (Figure \ref{fig:lee2}(b)). Now, $S_{11}$, which has the lowest workload among servers in CANDIDATES, is selected and invited to participate in the load distribution. However, $S_{11}$ rejects the invitation, because it is involved in another distribution, initiated by $S_{12}$. Thus, $S_{11}$ is removed from CANDIDATES and $S_{10}$ is selected because it now has the lowest workload among all servers in CANDIDATES (Figure \ref{fig:lee2}(c)). This process continues until the average workload is under a pre-defined threshold (Figure \ref{fig:lee2}(d) and Figure \ref{fig:lee2}(e)).

@@

%Um problema clï¿½ssico de alocaï¿½ï¿½o de tarefas em sistemas distribuï¿½dos ï¿½ o da dependï¿½ncia entre elas [TODO:refsDEdistGRAPH]. Tarefas dependentes entre si fazem com que os processadores nos quais elas estï¿½o sendo executadas tenham que se comunicar para que o processamento possa continuar. Isso gera dois problemas principais: em primeiro lugar, o processamento como um todo ï¿½ atrasado por causa do tempo de espera de cada transmissï¿½o e recebimento de mensagens -- supondo que a comunicaï¿½ï¿½o entre os processos seja por meio de mensagens, como ï¿½ o caso do MPI [TODO:ref].

@@

%Para resolver este problema, sï¿½o utilizados grafos, da seguinte maneira: o conjunto de tarefas a realizar ï¿½ mapeado em um grafo com pesos. Cada vï¿½rtice representa uma tarefa e cada aresta representa a comunicaï¿½ï¿½o entre as tarefas. O peso de cada vï¿½rtice representa o custo de processamento e o peso da aresta a carga de comunicaï¿½ï¿½o. Para fazer a distribuiï¿½ï¿½o, ï¿½ feito um particionamento do grafo gerado. Cada partiï¿½ï¿½o terï¿½ um conjunto de tarefas que serï¿½o executadas no mesmo nodo e, conseqï¿½entemente, o atraso de comunicaï¿½ï¿½o entre elas serï¿½ relativamente pequeno [TODO:refbli]. No entanto, as arestas do grafo que ligam partiï¿½ï¿½es diferentes representam comunicaï¿½ï¿½o entre diferentes nodos. A Figura [TODO:ref] ilustra essa situaï¿½ï¿½o.

@@

%[TODO:figura do grafo de tarefas e dependï¿½ncias, assim como particionamento com corte de arestas e diferentes particionamentos levando a diferentes cortes de aresta]

@@

%Os algoritmos de particionamento do grafo de tarefas, geralmente, buscam atingir dois objetivos: gerar partiï¿½ï¿½es de peso aproximadamente igual, ou seja, cada nodo terï¿½ uma carga computacional semelhante ï¿½ dos outros; e minimizar o corte de aresta, ou seja, fazer com que a dependï¿½ncia entre os conjuntos de tarefas seja a menor possï¿½vel, reduzindo a comunicaï¿½ï¿½o entre os nodos de processamento. Por corte de aresta, entende-se o somatï¿½rio dos pesos de todas as arestas que ligam vï¿½rtices que estï¿½o em partiï¿½ï¿½es diferentes. [TODO:reffig]

@@

%%TODO:ver esse "definiï¿½ï¿½es e parametros"
%Nas prï¿½ximas seï¿½ï¿½es, serï¿½ apresentada a soluï¿½ï¿½o proposta neste trabalho para balanceamento de carga, comeï¿½ando pelas definiï¿½ï¿½es e parï¿½metros utilizados, assim como serï¿½ descrito como o problema foi mapeado para grafos e quais os objetivos e critï¿½rios do esquema proposto.

@@

	\item \textbf{System usage}: fraction that indicates how much resources of the system as a whole is being used. It is defined by:
